---
title: "ZZSC9020 - Data Exploration"
author: "Zoren Liu"
date: "29 November 2022"
output:
  html_document:
    df_print: paged
  pdf_document: default
  html_notebook: default
---

Loading libraries.

```{r results = FALSE, warning=FALSE, message=FALSE}

library(GGally)
library(dplyr)
library(lubridate)

```

Load datasets. Note, `DATETIME` is read as class string, so it is converted to class `DATETIME`.

```{r}

temperature <- read.csv("temperature_nsw.csv")#, stringsAsFactors = F)
temperature$DATETIME <- as.POSIXct(temperature$DATETIME,format="%Y-%m-%d %H:%M:%S")

demand <- read.csv("totaldemand_nsw.csv")
demand$DATETIME <- as.POSIXct(demand$DATETIME,format="%Y-%m-%d %H:%M:%S")


forecast <- read.csv("forecastdemand_nsw.csv")
forecast$DATETIME <- as.POSIXct(forecast$DATETIME,format="%Y-%m-%d %H:%M:%S")

```

Let us view the data.

```{r}

str(temperature)
summary(temperature)

str(demand)
summary(demand)

str(forecast)
summary(forecast)

```

We notice the erroneous temperature measurements of -9999. We replace those values with the average of two adjacent temperature
measurements.

```{r}

ols <- which(temperature$TEMPERATURE < -50)
for (i in ols) {
  temperature$TEMPERATURE[i] <- mean(temperature$TEMPERATURE[i-1], temperature$TEMPERATURE[i+1])
}

```

We notice `demand` have 144 datetimes that are NA. We update them by refering to adjacent datetime records.

```{r}

na_times <- which(is.na(demand$DATETIME))

for (t in na_times) {
  demand$DATETIME[t] <- demand$DATETIME[t-1] + 5*60
}

```


We plot the distribution.

```{r fig.width = 12, fig.height = 10, fig.align = "center", message=FALSE}

plot(temperature$DATETIME, temperature$TEMPERATURE)

```

```{r fig.width = 12, fig.height = 10, fig.align = "center", message=FALSE}

plot(demand$DATETIME, demand$TOTALDEMAND)

```

Let us find the daily peak demand time.

```{r}

max_demand <- demand %>%
  mutate(by_date = floor_date(DATETIME, "day")) %>%
    group_by(by_date) %>%
    summarise(MAXDEMAND = TOTALDEMAND[which.max(TOTALDEMAND)], DATETIME = DATETIME, TOTALDEMAND=TOTALDEMAND) %>%
  filter(MAXDEMAND == TOTALDEMAND) %>%
  mutate(by_hour = format(as.POSIXct(DATETIME),   # Extract hours, minutes & seconds
                 format = "%H:%M:%S"))


# To plot time on y-axis, need a date to the time. Hence, added the same date too all `by_hour`.

max_demand$by_hour <- as.POSIXct(paste(max_demand$by_date[1], max_demand$by_hour), format="%Y-%m-%d %H:%M:%S")




```

```{r fig.width = 12, fig.height = 10, fig.align = "center", message=FALSE}

plot(max_demand$by_date, max_demand$by_hour)

```

A consistent trend we see is that the peak demand typically happens during the evening (appox 6pm - 8pm). This
is because everyone is home. In the first half of 2010s, we see trend where, during summer and winter times, the peak demand
are spread out from late morning to the evening (approx 11am - 6pm), most likely for cooling and heating. However, this 
seasonal variation in power demand reduces as time goes on. This might be because of the increase in roof top solar, where
the solar panels have taken over some of the demand from the grid.



